Sampling Strategies in Decisions from Experience

### 2022-01-31 manuscript.Rmd
Update introduction

<!-- Introduction -->

The mind is a cognitive system that operates on inputs from an environment or memory.
In the domain of decision making, such inputs can be organized in a set $I$ of choice alternatives $i \in I$, the latter being distinguishable from each other on the basis of the *properties* they possess.
The basic structure of decision making may then be modeled by the inclusion map

@@

[cf. @jostMathematicalConcepts2015, pp. 14-15], indicating that decision making is a *process* that starts from an input set $I$, including all choice alternatives, and stops at a proper subset $I' \subset I$, including only the chosen alternatives. 
Given that the choice is perfectly systematic---as opposed to random or noisy---the chosen alternatives in $I'$ should share some property or combination of several properties which the other alternatives in the complementary subset $I \setminus I'$ do not possess.
In the following, I will refer to statements about properties that serve to distinguish between the chosen and non-chosen alternatives as *selection criteria*.

>>>>

<!--The mind is a cognitive system that operates on inputs from an environment or memory.
In the domain of decision making, such inputs can be organized in a set $I$ of choice alternatives $i \in I$, the latter being distinguishable from each other on the basis of the *properties* they possess.
The basic structure of decision making may then be modeled by the inclusion map

@@

[cf. @jostMathematicalConcepts2015, pp. 14-15], indicating that decision making is a process that starts from an input set $I$, including all choice alternatives, and stops at a proper subset $I' \subset I$, including only the chosen alternatives [see @heOntologyDecisionModels2020].-->

Since there is an uncountable number of decisions people confront in the natural world, behavioral decision research routinely abstracts from particular decisions, e.g., the choice between job offers, political parties to vote for, travel destinations, investment plans, and whatnot.
Rather, with the choice between at least two *prospects* (or: gambles), it studies a case which is counterfactual in that it omits the many particularities of each decision and the differences between them, but retains in the form of prospects the fundamental properties that almost all choice alternatives are assumed to possess [@lopesThoughtsPsychologicalConcept1983].^[See Chapter 1 in @knightRiskUncertaintyProfit1921, for a discussion of the reasonableness of such abstractions.] 
These fundamental properties are considered to be the possible outcomes of a choice alternative---hereafter denoted by the Greek letter $\omega$---and the probabilities with which these outcomes occur following the choice of the alternative---hereafter $p(\omega)$.
Thus, prospects are abstract surrogates for choice alternatives in the natural world, resembling probability mass functions of the form 
$$
p: \Omega \mapsto [0,1]
\; ,
\tag{1}
$$ 
each of which assigns the possible outcomes in a finite set $\Omega = \{\omega_1, ..., \omega_n\}$ a probability $0 < p(\omega_i) \leq 1$, where $i \in \{1, ..., n\}$ and $\sum_i^n p(\omega_i) = 1$.

The choices people actually make between such prospects are often described in terms of their deviation from the principle of expected value (EV) maximization, where the EV is given by 
$$
EV = \sum_i^n \omega_i \times p(\omega_i)
\; .
\tag{2}
$$ 
That is, following the principle of EV maximization, people would first calculate the prospects' EV according to (2) and thereafter chose the prospect for which the EV is maximized.
<!--However, the many assessments of peoples' choices in these decisions neither corroborate the theory that people follow the principle of EV maximization nor that of expected utility (EU) maximization [e.g., @kahnemanProspectTheoryAnalysis1979, ...].^[Following the EU principle [@bernoulliExpositionNewTheory1954], the objective outcomes in (2) are transformed by a concave utility function. This transformation expresses the idea that an increase in the objective outcome yields a diminishing increase in its subjective utility, where the latter is weighted by the objective probability to compute the expected utility of a prospect. Note that for negative outcomes the transformation function is convex.]-->
To *describe* peoples' deviations from EV <!--and EU--> maximization [see @erevAnomaliesForecastsDescriptive2017, for a recent replication of classical demonstrations], the models of cumulative prospect theory [CPT, @tverskyAdvancesProspectTheory1992; see also @stottCumulativeProspectTheory2006, for an overview of models] may replace the objective outcomes and probabilities in (2) by nonlinear transformations thereof.
More specifically, the models contain a combination of two parameterized <!--monotonic--> functions, one transforming the objective outcomes---referred to as the *value function*, $\omega \mapsto v(\omega)$---and the other transforming the cumulative distribution function of objective probabilities---referred to as the *probability weighting function*, $p(\omega) \mapsto w(p(\omega))$ or $p(\omega) \mapsto w(p(\omega))$.
<!--Note that the probability weighting function is a composition of two functions $\omega \maps to p(\omega) \maps to w(p(\omega))$.-->

The shape of the probability weighting function lead to the common interpretation that people choose as if they were over- or underweighting outcomes of certain probability (see Figure 1). 
```{r}
wf <- tibble(gamma = c(.7, 1, 1.5), 
             type = c("overweighting", "linear", "underweighting")) %>%
  expand_grid(prob = seq(0, 1, .01)) %>%
  mutate(w = exp(-1*(-log(prob))^gamma)) %>% 
  select(-gamma)

ggplot(wf) +
  geom_path(aes(prob, w, linetype = type)) +
   scale_linetype_manual(values=c("solid", "dotted", "longdash")) +
  scale_x_continuous(breaks = seq(0, 1, .1)) +
  scale_y_continuous(breaks = seq(0, 1, .1)) + 
  labs(x = expression(paste("p(", omega, ")")),
       y = expression(paste("w(p(", omega, "))")),
       linetype = "Weighting Pattern") + 
  theme_apa() 
```

One such model is the cumulative prospect theory [@tverskyAdvancesProspectTheory1992]

$$
\tag{3}
$$

Given that the choice is perfectly systematic---as opposed to random or noisy---the chosen alternatives in $I'$ should share some property or combination of several properties which the other alternatives in the complementary subset $I \setminus I'$ do not possess.
In the following, I will refer to statements about properties that serve to distinguish between the chosen and non-chosen alternatives as *selection criteria*.
